Revealing Transient Strain in Geodetic Data with Gaussian Process Regression

### 2017-06-18 HinesHetlandGPR.tex
added more

We constrain $u$ with GNSS data, which records $u$ as well as other physical and non-physical processes which we are not interested in. We describe GNSS observations at position $\vec{x}_i$ and time $t_j$ as a realization of the random variable 
\begin{align}\label{eq:Data}
\begin{split}
d_{ij} = &u(\vec{x}_i,t_j) + \eta(\vec{x}_i,t_j) + w_{ij} + a^{(1)}_i + a^{(2)}_it_j + \\
         &a^{(3)}_i\sin(2 \pi t_j) + a^{(4)}_i\cos(2 \pi t_j) + a^{(5)}_i\sin(4 \pi t_j) + a^{(6)}_i\cos(4 \pi t_j), 
\end{split}
\end{align}
where $a^{(1)}_{i}$ is an offset that is unique for each GNSS monument, $a^{(2)}_{i}$ is the secular velocity at $\vec{x}_i$, and the sinusoids describe seasonal deformation (using units of years for $t_j$). We use $w_{ij}$ to denote normally distributed, uncorrelated noise. Correlated noise which does not have a parametric representation is denoted by $\eta$.  For example, $\eta$ can consist of temporally correlated noise describing benchmark wobble \citep[e.g.,][]{Wyatt1982,Wyatt1989}, and/or spatially correlated noise describing common mode error \citep[e.g.,][]{Wdowinski1997}. For now, we will only assume that $\eta \sim \mathcal{N}(0,C_\eta)$. We consider the six coefficients in eq. (\ref{eq:Data}) to be uncorrelated random variables distributed as $\mathcal{N}(0,\kappa^2)$ in the limit as $\kappa \to \infty$ (i.e., the coefficients have diffuse priors). Of course, the secular velocities, $a^{(2)}_{i}$, are spatially correlated and we could invoke a tectonic model to form a prior on $a^{(2)}_{i}$. However, in our application to Cascadia, we will be using displacement time series which are long enough to sufficiently constrain $a^{(2)}_{i}$ for each station, avoiding the need to incorporate a prior on it. Likewise, the seasonal coefficients may be spatially correlated \citep{Langbein2008}, and it may be worth exploring and exploiting such a correlation in a future study.

@@

\section{Discussion}\label{sec:Discussion}
We have demonstrated that transient strain rates estimated with the method described in Section \ref{sec:Method} can be used to detect SSEs, and will be robust to detect other transient geophysical phenomena. Another potential application would be to use the GNSS derived transient strain rates to develop noise models for borehole strain meters (BSMs). The plate Boundary Observatory maintains 43 BSMs in the Pacific Northwest, and it has been demonstrated that BSMs are able to record transient geophysical events such as SSEs \citep[e.g.,][]{Dragert2011}. However, noise in BSM data is not well understood, which complicates the use of BSM data for identifying geophysical signal. The noise consists, in part, of a long-term decay resulting from the instrument equilibrating with the surrounding rock \citep{Gladwin1987}. Typically, this noise is dealt with in an ad-hoc manner by fitting and removing exponentials and low-order polynomials. A more rigorous quantification of BSM noise could be performed by using GNSS derived strain rates as an approximation of the signal in BSM data.

We also note that there is potential for a more thorough analysis of the spatio-temporal noise in GNSS data, $v$, than what was performed in Section \ref{sec:NoiseModel}. In this study, we have not explored the spatial covariance of $v$, since it would not have significantly affected our estimated strain rates. If, for example, we were interested in inferring fault slip from GNSS displacements then it may be necessary to account for the spatial covariance of the $v$.

In future studies, we may explore more sophisticated seasonal models than what is used in eq. (\ref{eq:Data}). As suggested earlier, we may want to quantify the spatial covariance of seasonal deformation. We may also want to explore whether the semiannual terms in eq. (\ref{eq:Data}) are necessary or if higher frequency terms are warranted.  The periodic Gaussian process \citep{Mackay1998} is an alternative model for seasonal deformation and is well suited for exploring the roughness of seasonal deformation.  The periodic Gaussian process has zero mean and covariance function,
\begin{equation}\label{eq:Periodic}
T(t,t') = \phi \exp\left(\frac{-\sin(\pi|t - t'|)^2}{2\tau^2}\right).
\end{equation}
Realizations have annual periodicity and the roughness is controlled by $\tau$. Decreasing $\tau$ has the same effect as including higher frequency sinusoids in the seasonal model. The optimal value for $\tau$ can be found with the REML method as described in Section \ref{sec:NoiseModel}.

>>>>

We constrain $u$ with GNSS data, which records $u$ as well as other physical and non-physical processes which we are not interested in. We describe GNSS observations at position $\vec{x}_i$ and time $t_j$ as a realization of the random variable 
\begin{align}\label{eq:Data}
\begin{split}
d_{ij} = &u(\vec{x}_i,t_j) + \eta(\vec{x}_i,t_j) + w_{ij} + a^{(1)}_i + a^{(2)}_it_j + \\
         &a^{(3)}_i\sin(2 \pi t_j) + a^{(4)}_i\cos(2 \pi t_j) + a^{(5)}_i\sin(4 \pi t_j) + a^{(6)}_i\cos(4 \pi t_j), 
\end{split}
\end{align}
where $a^{(1)}_{i}$ is an offset that is unique for each GNSS monument, $a^{(2)}_{i}$ is the secular velocity at $\vec{x}_i$, and the sinusoids describe seasonal deformation (using units of years for $t_j$). We use $w_{ij}$ to denote normally distributed, uncorrelated noise. Correlated noise which does not have a parametric representation is denoted by $\eta$.  For example, $\eta$ can consist of temporally correlated noise describing benchmark wobble \citep[e.g.,][]{Wyatt1982,Wyatt1989}, and/or spatially correlated noise describing common mode error \citep[e.g.,][]{Wdowinski1997}. For now, we will only assume that $\eta \sim \mathcal{N}(0,C_\eta)$. We consider the six coefficients in eq. (\ref{eq:Data}) to be uncorrelated random variables distributed as $\mathcal{N}(0,\kappa^2)$ in the limit as $\kappa \to \infty$ (i.e., the coefficients have diffuse priors). Of course, the secular velocities, $a^{(2)}_{i}$, are spatially correlated and we could invoke a tectonic model to form a prior on $a^{(2)}_{i}$. However, in our application to Cascadia, we will be using displacement time series which are long enough to sufficiently constrain $a^{(2)}_{i}$ for each station, avoiding the need to incorporate a prior on it. Likewise, seasonal deformation is spatially correlated \citep{Dong2002,Langbein2008}, and it may be worth exploring and exploiting such a correlation in a future study.

@@

\section{Discussion}\label{sec:Discussion}
We have demonstrated that transient strain rates estimated with the method described in Section \ref{sec:Method} can be used to detect SSEs and will be robust to detect other transient geophysical phenomena. Another potential application would be to use GNSS derived transient strain rates to develop noise models for borehole strain meters (BSMs). The Plate Boundary Observatory maintains 43 BSMs in the Pacific Northwest, and it has been demonstrated that BSMs are able to record transient geophysical events such as SSEs \citep[e.g.,][]{Dragert2011}. However, noise in BSM data is not well understood, which complicates the use of BSM data for detecting geophysical events. The noise consists, in part, of a long-term decay resulting from the instrument equilibrating with the surrounding rock \citep{Gladwin1987}. Typically, this noise is dealt with in an ad-hoc manner by fitting and removing exponentials and low-order polynomials. A more rigorous quantification of BSM noise could be performed by analyzing the residuals between the the BSM data and GNSS derived strain rates.

There is potential for a more thorough analysis of the spatio-temporal noise in GNSS data, $\eta$, than what was performed in Section \ref{sec:NoiseModel}. In this study, we did not explored the spatial covariance of $\eta$. Spatially correlated common mode noise \citep{Wdowinski1997}, which results from factors such as reference frame error, is a non-negligible component of GNSS data. We ignore common mode error in this study because transients strain rates are insensitive to it. For other geophysical studies based on GNSS data, such as fault slip inversions, it may be necessary to incorporate a spatially covarying noise model \citep[e.g.,][]{Miyazaki2003}. We can also improve upon the season model used in this study, which consists of four spatially uncorrelated sinusoids for each station. We did not explore the spatial covariance of seasonal deformation or the temporal ``roughness'' (i.e., the number of sinusoids needed to describe the observations). The periodic Gaussian process \citep{Mackay1998} is an alternative model for seasonal deformation and is well suited for exploring the roughness of seasonal deformation.  The periodic Gaussian process has zero mean and covariance function,
\begin{equation}\label{eq:Periodic}
T(t,t') = \phi \exp\left(\frac{-\sin(\pi|t - t'|)^2}{2\tau^2}\right).
\end{equation}
Realizations have annual periodicity and the roughness is controlled by $\tau$. Decreasing $\tau$ has the same effect as including higher frequency sinusoids in the seasonal model. The optimal value for $\tau$ can be found with the REML method as described in Section \ref{sec:NoiseModel}.
